# Answer to Rev3 coloc comment

We want to see wether the increased heterogeneity in expression values in tumor contributes to the delta-eQTL effect.

```{r}
library(edgeR, help, pos = 2, lib.loc = NULL)

expr_values <- readRDS("/mnt/raid5_hdd/matsam01/MENDEL_LORD/processed_data/LORD_project/RNAseq/processed_data/kallisto/txi.rds")$counts
genes <- read.table("genes_for_DEG.txt")$V1
expr_values <- expr_values[genes, ]
```

## Make model

In the first step, we make a EdgeR model for gene expression.

```{r}
# 1. Make sure counts is genes × samples
# If your current matrix is samples × genes, transpose:
counts <- expr_values  # remove this line if it's already genes × samples

# 2. Extract pair and condition from sample names
sample_names <- colnames(counts)
# Example: "some_random_prefix_1_A", "some_random_prefix_1_B", etc.
split_names <- strsplit(sample_names, "_")

pair    <- sapply(split_names, function(s) s[length(s) - 1])
condition <- sapply(split_names, function(s) s[length(s)])

# Turn into factors (order matters if you want specific baselines)
pair      <- factor(pair)
condition <- factor(condition, levels = c("Sain", "Tumeur"))
levels(condition)[levels(condition) == "Sain"]   <- "Lung"
levels(condition)[levels(condition) == "Tumeur"] <- "Tumor"
```

```{r}
# 3. Build DGEList and pipeline
y <- DGEList(counts = counts)
y <- calcNormFactors(y)

# 4. Design matrix: pair as blocking factor, condition of interest
design <- model.matrix(~ condition)
colnames(design)
```

```{r}
# 5. Estimate dispersion and fit NB GLM
y <- estimateDisp(y, design)
fit <- glmFit(y, design)
```

```{r}
# 6. Extract residuals
fitted_vals <- fit$fitted.values  # genes × samples, on count scale

# Deviance residuals: sign(observed - fitted) * sqrt(d_i)
# EdgeR stores the deviance components; easiest to compute directly:
obs <- fit$counts
mu  <- fitted_vals
phi <- fit$dispersion

# Poisson-style deviance residuals (approximation)
dev_res <- sign(obs - mu) * sqrt(2 * (ifelse(obs == 0, 0, obs * log(obs / mu)) - (obs - mu)))

# pearson residuals
var_nb  <- mu + phi * mu^2  
pear_res <- (obs - mu) / sqrt(var_nb)
```

## Validation of increased heterogeneity

Second, we need to validate that tumors indeed have an increased heterogeneity. We do that by comparing the deviance residuals of the lung and tumor samples.

```{r}
# Split deviance residuals by condition
# pear_res is genes × samples; we want per-gene distributions
# Melt into long format for plotting and testing
sain_idx   <- which(condition == "Lung")
tumeur_idx <- which(condition == "Tumor")

# Per-sample median residual (one value per gene)
sain_medians   <- apply(pear_res[, sain_idx,   drop = FALSE], 2, median)
tumeur_medians <- apply(pear_res[, tumeur_idx, drop = FALSE], 2, median)

# Sign test: for each pair, is Tumor > Lung?
# Assumes paired samples (same pair IDs, matched by pair factor)

diffs <- sapply(genes, function(g) {
    t_idx <- which(condition == "Tumor")
    s_idx <- which(condition == "Lung")
    mean(abs(pear_res[g, t_idx])) - mean(abs(pear_res[g, s_idx]))
})

lung_dev <- sapply(genes, function(g) {
    t_idx <- which(condition == "Tumor")
    s_idx <- which(condition == "Lung")
    mean(abs(pear_res[g, s_idx]))
})

tumor_dev <- sapply(genes, function(g) {
    t_idx <- which(condition == "Tumor")
    s_idx <- which(condition == "Lung")
    mean(abs(pear_res[g, t_idx]))
})

n_pos    <- sum(diffs > 0)
n_nonzero <- sum(diffs != 0)

# Exact binomial sign test (H0: median diff = 0, two-sided)
sign_test <- binom.test(n_pos, n_nonzero, p = 0.5, alternative = "two.sided")

cat("=== Sign Test: Deviance Residuals (Tumor vs Lung) ===\n")
cat(sprintf("genes with absolute Tumor deviance > Lung deviance : %d / %d\n", n_pos, n_nonzero))
cat(sprintf("p-value: %.4e\n", sign_test$p.value))
```

```{r}
library(ggplot2)

df <- data.frame(
    deviance = c(lung_dev, tumor_dev),
    condition = factor(c(rep("Lung", length(lung_dev)), 
                       rep("Tumor", length(tumor_dev))),
                     levels = c("Lung", "Tumor"))
)

ggplot(df, aes(x = condition, y = deviance, fill = condition)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Lung" = "#ffbf0086", "Tumor" = "#1a004aa3")) +
  labs(x = "Condition", y = "Mean absolute pearson residuals (per gene)") +
  theme_classic(base_size = 28) +
  theme(
    legend.position = "none",
    aspect.ratio = 1
  )
ggsave("img/pearson_residuals_lung_vs_tumor.pdf", width = 10, height = 10)
```

## Any relationship between expression levels and mean/mdeian deviance?

load the abundance values

```{r}
abundance_values <- readRDS("/mnt/raid5_hdd/matsam01/MENDEL_LORD/processed_data/LORD_project/RNAseq/processed_data/kallisto/txi.rds")$abundance
abundance_values <- abundance_values[rownames(fit$counts), colnames(fit$counts)]
```

```{r}
mean_expr_sain   <- rowMeans(log1p(abundance_values[, sain_idx]))
mean_expr_tumeur <- rowMeans(log1p(abundance_values[, tumeur_idx]))

df_scatter <- data.frame(
  deviance   = c(lung_dev, tumor_dev),
  expression = c(mean_expr_sain, mean_expr_tumeur),
  condition  = factor(c(rep("Lung", length(lung_dev)),
                        rep("Tumor", length(tumor_dev))),
                      levels = c("Lung", "Tumor"))
)

x_lo <- quantile(df_scatter$expression, 0.01)
x_hi <- quantile(df_scatter$expression, 0.99)

ggplot(df_scatter, aes(x = expression, y = deviance, color = condition)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "loess", se = TRUE) +
  scale_color_manual(values = c("Lung" = "#ffbf0086", "Tumor" = "#1a004aa3")) +
  labs(x = "Mean log1p expression", y = "Mean absolute pearson residuals (per gene)",
       color = "Condition") +
  theme_classic(base_size = 28) +
  coord_cartesian(xlim = c(x_lo, x_hi)) +
  theme(aspect.ratio = 1)
  ggsave("img/pearson_residuals_vs_gene_expr.pdf", width = 10, height = 10)
```

## Any relationship between mean/median devicance change and Delta-eQTL effect size?

```{r}
delta_eqtl <- read.csv("/mnt/raid5_hdd/matsam01/genetic_association_sumstats/lung_expression_LORD2024/Delta_eQTLs_60_peer_best_hit.tsv", sep = "\t")
```

```{r}
# Build a named vector from diffs (assuming names are gene IDs)
diffs_df <- data.frame(
  phe_id = names(diffs),
  diffs  = as.numeric(diffs),
  tumor = tumor_dev,
  lung = lung_dev
)

# Merge with delta_eqtl by gene ID
df_eqtl <- merge(diffs_df, delta_eqtl[, c("phe_id", "slope", "slope_lung", "slope_tumor")], by = "phe_id")

x_lo <- quantile(df_eqtl$slope, 0.001)
x_hi <- quantile(df_eqtl$slope, 0.999)

ggplot(df_eqtl, aes(x = abs(slope), y = diffs)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(data = subset(df_eqtl, slope >= x_lo & slope <= x_hi),
              method = "loess", se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  labs(x = "delta-eQTL slope",
  y = "Difference of mean absolute pearson residuals") +
  theme_classic(base_size = 28) +
  theme(aspect.ratio = 1)
  ggsave("img/pearson_residuals_vs_delta_effect_size.pdf", width = 10, height = 10)
```

## Any relationship between eQTL effect size fold change / difference and Deviance difference?

```{r}
df_eqtl$slope_diff <- df_eqtl$slope_tumor - df_eqtl$slope_lung

x_lo <- quantile(df_eqtl$slope_diff, 0.001)
x_hi <- quantile(df_eqtl$slope_diff, 0.999)

ggplot(df_eqtl, aes(x = abs(slope_diff), y = diffs)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(data = subset(df_eqtl, slope_diff >= x_lo & slope_diff <= x_hi),
              method = "loess", se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  labs(x = "Absolute eQTL effect size difference",
  y = "Difference of mean absolute pearson residuals") +
  theme_classic(base_size = 28) +
  theme(aspect.ratio = 1)
  ggsave("img/pearson_residuals_vs_tum-lung_effect_sizes.pdf", width = 10, height = 10)
```
